---
title: 'Mammal Diversity: Mammalogy 2018, Hypothesis testing'
author: "Erika"
date: "12/04/2018"
output: html_document
---
## Introduction
The students in Mammalogy are working on analyzing their mammal trapping data (including game camera data) for their final report.  Here is the work that needs to take place.

On problem set 2, they were asked to only deal with the small mammal data.  

This work requires two files, one called "Kip_trapping_data.csv" and the other called "SpeciesByTransect"

## Set up the workspace
Set some basic parameters 
```{r}
rm(list =ls()) #clear's R workspace
library(dplyr)
library(ggplot2)
library(readr)

```

## Load and tidy some data
Load the new tree data
```{r}
Small <- read_csv("Kip_trapping_data.csv")
Camera<-read_csv("SpeciesByTransect")
```
Now clean up some of the small mammal data by getting rid of columns I don't want to see

```{r}
Wrong<-c(12:17)
Small<-Small[,-Wrong]
```
Now summarize the small mammal data for ease of use

```{r}
Small_spec_by_trans<-Small %>% group_by(Habitat, Transect, Species) %>% summarize(
	number = length(Species),
)
Small_by_habitat<-Small %>% group_by(Habitat, Species) %>% summarize(
	number = length(Species)
)
write.csv(Small_spec_by_trans, "smallmammalresults.csv") #this is the data set I need to work with for hyptothesis testing for PS2.  
```
Now my 2 summary data frames are about the same.  Let's get them formatted correctly

```{r}
Camera<-Camera[,-1]
#now change column names
colnames(Camera)<-c("Habitat", "Transect", "Species","Number")
colnames(Small_spec_by_trans)<-c("Habitat", "Transect", "Species","Number")
```
Now for compatibility I need to change the species names in Camera to match those in Small...

```{r}
small_species<-as.factor(Small_spec_by_trans$Species)
camera_species<-as.factor(Camera$Species)
#now get rid of non-mammals in Camera
#"BLUEJAY"          "CROWORRAVEN"      "DEERWHITETAILED"  "HUMAN"           
#[5] "MOLE"             "OPOSSUM"          "OTHERBIRD"        "OTHERSMALLMAMMAL"
# [9] "PORCUPINE"        "RACCOON"          "SQUIRRELFLYING"   "SQUIRRELGRAY"    
#[13] "SQUIRRELRED"      "TURKEY"           "WEASEL"          
notamammal<-c("BLUEJAY","CROWORRAVEN", "OTHERBIRD", "TURKEY")
```
Now select the species that are not mammals
```{r}
nonmammals<-which(Camera$Species =="BLUEJAY"|Camera$Species =="CROWORRAVEN"|Camera$Species =="OTHERBIRD"|Camera$Species =="TURKEY")
```
Now remove them from the dataframe
```{r}
Mammals<-Camera[-nonmammals,]
```

Now change species names

```{r}
Possums<-which(Mammals$Species =="OPOSSUM")
Mammals$Species[Possums]<-"Didelphis virginiana"
Humans<-which(Mammals$Species == "HUMAN")
#this worked; sometimes the levels aren't correctly recorded
Mammals$Species[Humans]<-"Homo sapiens"
Deer<-which(Mammals$Species =="DEERWHITETAILED")
Mammals$Species[Deer]<-"Odocoileus virginianus"
Mole<-which(Mammals$Species == "MOLE")
Mammals$Species[Mole]<- "Talpid"
Other<-which(Mammals$Species == "OTHERSMALLMAMMAL")
Mammals$Species[Other]<-"Other small mammal"
Porkies<-which(Mammals$Species =="PORCUPINE")
Mammals$Species[Porkies]<-"Erethizon dorsatum"
Raccoon<-which(Mammals$Species =="RACCOON")
Mammals$Species[Raccoon]<-"Procyon lotor"
Volans<-which(Mammals$Species =="SQUIRRELFLYING")
Mammals$Species[Volans]<-"Glaucomys volans"
grays<-which(Mammals$Species =="SQUIRRELGRAY")
Mammals$Species[grays]<-"Sciurus carolinensis"
reds<-which(Mammals$Species =="SQUIRRELRED")
Mammals$Species[reds]<-"Tamiasciurus hudsonicus"
Mustela<-which(Mammals$Species =="WEASEL")
Mammals$Species[Mustela]<-"Mustela sp."
```
Now the two df's match in structure.  I think we can now bind them together.

```{r}
Mammals$Number<-as.integer(Mammals$Number)
Mammals<-as.data.frame(Mammals)
Small_spec_by_trans<-as.data.frame(Small_spec_by_trans)
AllData<-rbind(Mammals,Small_spec_by_trans)
```
Now rename some things and clean things up
```{r}
rm(Small)
rm(Camera)
Small<-Small_spec_by_trans
rm(Small_spec_by_trans)
```
## Hypothesis Testing

* Test the hypothesis that the mean Species richness is different among the three forest habitats.  (Hint: What is your number of replicates for each habitat?)
* Test the hypothesis that the mean Simpson's index is different among the three forest habitats.  (Hint: What is your number of replicates for each habitat?)


###Small mammal data first
For the first hypothesis, I need to run an ANOVA with species richness as the response variable and habitat as the treatment.  There are 3 habitats, each with two replicates, so the power to detect differences is not great.

First I need to take the data by habitat and transect to get the number of species in each transect/habitat
```{r}
Small_midstep<-Small %>% group_by(Habitat, Transect, Species) %>% summarize(
	number = length(Species)
)
#now get SR by habitat/transect
Small_forGLM<-Small_midstep %>% group_by(Habitat, Transect) %>% summarize(
	number = sum(number)
)

```
Now run the glm with the dataset I just generated
```{r}
SmallLM<-lm(number~Habitat, data= Small_forGLM)
summary(SmallLM)
anova(SmallLM)
#now get tukey's least sig diff
#first determine the test statistic value with 2 and 3 df
qt(0.95, 2, 3)
SmallTukey<-TukeyHSD(aov(number~Habitat, Small_forGLM))

```
Now I need to do same, but use Simpson's index rather than SR
Start by reading in the simpson's data generated from PS2_SpecRich_Simpson.Rmd
```{r}
Simpson<-read_csv("smallMammalSimpsons.csv")
```
Now I can run the analysis

Let's look at the means in Simpson's index

```{r}
SmallSimp<-lm(Simpson~Habitat, data= Simpson)
summary(SmallSimp)
anova(SmallSimp)
#now get tukey's least sig diff
#first determine the test statistic value with 2 and 3 df
TukeyHSD(aov(Simpson~Habitat, Simpson))
```

###Game camera data

Now do the same analysis for the game camera data. However, for the Hemlock habitat, we only had game camera data from one of the two transects, so we cannot get a mean SR or Simpson's index for the game camera data alone.  No reason to continue.

###All data
And now for all of the data (game camera + small mammal)

First I need the data. 
```{r}
All<-read_csv("AllMammSR.csv")
```

Now run the glm with the dataset I just generated
```{r}
AllLM<-lm(SR~Habitat, data= All)
summary(AllLM)
anova(AllLM)
#difference is NS so no need to run Tukeys.
```
Now I need to do same, but use Simpson's index rather than SR
Start by reading in the simpson's data 
```{r}
AllSimpson<-read_csv("AllMammalsSimpsons.csv")
```
Now I can run the analysis

Let's look at the means in Simpson's index

```{r}
AllSimp<-lm(Simpson~Habitat, data= AllSimpson)
summary(AllSimp)
anova(AllSimp)
#no sig dif (P = 0.78) so no need to run Tukey's
```


