---
title: 'Mammal Diversity: Mammalogy 2018'
author: "Erika"
date: "12/04/2018"
output: html_document
---
## Introduction
The students in Mammalogy are working on analyzing their mammal trapping data (including game camera data) for their final report.  Here is the work that needs to take place.

This work requires two files, one called "Kip_trapping_data.csv" and the other called "SpeciesByTransect"

## Set up the workspace
Set some basic parameters 
```{r}
rm(list =ls()) #clear's R workspace
library(dplyr)
library(ggplot2)
library(readr)
```

###Step 1. Mammal diversity
Students are asked to calculate average species richness and average Simpson's index for each habitat, using each of the 2 transects within each habitat as a replicate. They are also to provide the standard error of the mean.

#### Load some data
Load the new tree data
```{r}
Small <- read_csv("Kip_trapping_data.csv")
Camera<-read_csv("SpeciesByTransect")
```
Now clean up some of the small mammal data by getting rid of columns I don't want to see

```{r}
Wrong<-c(12:17)
Small<-Small[,-Wrong]
```
Now summarize the small mammal data for ease of use

```{r}
Small_spec_by_trans<-Small %>% group_by(Habitat, Transect, Species) %>% summarize(
	number = length(Species),
)
Small_by_habitat<-Small %>% group_by(Habitat, Species) %>% summarize(
	number = length(Species)
)
write.csv(Small_spec_by_trans, "smallmammalresults.csv")
```
Now my 2 summary data frames are about the same.  Let's get them formatted correctly

```{r}
Camera<-Camera[,-1]
#now change column names
colnames(Camera)<-c("Habitat", "Transect", "Species","Number")
colnames(Small_spec_by_trans)<-c("Habitat", "Transect", "Species","Number")
```
Now for compatibility I need to change the species names in Camera to match those in Small...

```{r}
small_species<-as.factor(Small_spec_by_trans$Species)
camera_species<-as.factor(Camera$Species)
#now get rid of non-mammals in Camera
#"BLUEJAY"          "CROWORRAVEN"      "DEERWHITETAILED"  "HUMAN"           
#[5] "MOLE"             "OPOSSUM"          "OTHERBIRD"        "OTHERSMALLMAMMAL"
# [9] "PORCUPINE"        "RACCOON"          "SQUIRRELFLYING"   "SQUIRRELGRAY"    
#[13] "SQUIRRELRED"      "TURKEY"           "WEASEL"          
notamammal<-c("BLUEJAY","CROWORRAVEN", "OTHERBIRD", "TURKEY")
```
Now select the species that are not mammals
```{r}
nonmammals<-which(Camera$Species =="BLUEJAY"|Camera$Species =="CROWORRAVEN"|Camera$Species =="OTHERBIRD"|Camera$Species =="TURKEY")
```
Now remove them from the dataframe
```{r}
Mammals<-Camera[-nonmammals,]
```

Now change species names

```{r}
Possums<-which(Mammals$Species =="OPOSSUM")
Mammals$Species[Possums]<-"Didelphis virginiana"
Humans<-which(Mammals$Species == "HUMAN")
#this worked; sometimes the levels aren't correctly recorded
Mammals$Species[Humans]<-"Homo sapiens"
Deer<-which(Mammals$Species =="DEERWHITETAILED")
Mammals$Species[Deer]<-"Odocoileus virginianus"
Mole<-which(Mammals$Species == "MOLE")
Mammals$Species[Mole]<- "Talpid"
Other<-which(Mammals$Species == "OTHERSMALLMAMMAL")
Mammals$Species[Other]<-"Other small mammal"
Porkies<-which(Mammals$Species =="PORCUPINE")
Mammals$Species[Porkies]<-"Erethizon dorsatum"
Raccoon<-which(Mammals$Species =="RACCOON")
Mammals$Species[Raccoon]<-"Procyon lotor"
Volans<-which(Mammals$Species =="SQUIRRELFLYING")
Mammals$Species[Volans]<-"Glaucomys volans"
grays<-which(Mammals$Species =="SQUIRRELGRAY")
Mammals$Species[grays]<-"Sciurus carolinensis"
reds<-which(Mammals$Species =="SQUIRRELRED")
Mammals$Species[reds]<-"Tamiasciurus hudsonicus"
Mustela<-which(Mammals$Species =="WEASEL")
Mammals$Species[Mustela]<-"Mustela sp."
```
Now the two df's match in structure.  I think we can now bind them together.

```{r}
Mammals$Number<-as.integer(Mammals$Number)
Mammals<-as.data.frame(Mammals)
Small_spec_by_trans<-as.data.frame(Small_spec_by_trans)
AllData<-rbind(Mammals,Small_spec_by_trans)
```
Now rename some things and clean things up
```{r}
rm(Small)
rm(Camera)
Small<-Small_spec_by_trans
rm(Small_spec_by_trans)
```
#### Summary analysis
1. Create a table, including a complete table heading, that describes the species of mammal found in the Kip Tract in total and on each of the sub-habitats. Include in your table the number trapped for each species in each habitat and in total

#####Small mammal data first
```{r}
SmallTab1<-Small%>%group_by(Habitat, Species) %>% summarise(
	total=sum(Number, na.rm = T)
)

SmallTotal<-SmallTab1 %>% group_by(Species) %>% summarise(
	tot = sum(total)
)
```
#####Game camera data
```{r}
GameTab1<-Mammals%>%group_by(Habitat, Species) %>% summarise(
	total=sum(Number, na.rm = T)
)

GameTotal<-GameTab1 %>% group_by(Species) %>% summarise(
	tot = sum(total)
)
```
#####All data
```{r}

AllTab1<-AllData%>%group_by(Habitat, Species) %>% summarise(
	total=sum(Number, na.rm = T)
)

AllTotal<-AllTab1 %>% group_by(Species) %>% summarise(
	tot = sum(total)
)
```

#### Species Richness
2a.What is species richness of mammals on the Kip Tract altogether without considering sub-habitats? 
2b. What is the mean species richness with SEM in each habitat?
#####Small mammals 
```{r}
#2a.
SRall<-n_distinct(SmallTab1$Species)#species richness of all small mammals
#what are those species?
SpeciesSmall<-levels(SmallTab1$Species)
#2b.
SRbyHabSmall<-Small %>% group_by(Habitat, Transect) %>% summarise(
	SR = n_distinct(Species)
)
#now get mean, sd
SRdetailsSmall<-SRbyHabSmall%>% group_by(Habitat) %>% summarise(
	meanSR = mean(SR, na.rm = T),
	semSR = sd(SR, na.rm = T)/sqrt(sum(!is.na(SR)))
)
```
#####Game cameras
```{r}
#2a.
SRall<-n_distinct(GameTab1$Species)#species richness of all small mammals
#what are those species?
SpeciesGame<-levels(GameTab1$Species)
#2b.
SRbyHabGame<-Mammals %>% group_by(Habitat, Transect) %>% summarise(
	SR = n_distinct(Species)
)
#now get mean, sd
SRdetailsGame<-SRbyHabGame%>% group_by(Habitat) %>% summarise(
	meanSR = mean(SR, na.rm = T),
	semSR = sd(SR, na.rm = T)/sqrt(sum(!is.na(SR)))
)
#now save summary

write.csv(SRbyHabGame, "GameCamSRsummary.csv")
```

#####All data
```{r}
#2a.
SRall<-n_distinct(AllTab1$Species)#species richness of all small mammals
#what are those species?
SpeciesAll<-levels(AllTab1$Species)
#2b.
SRbyHabAll<-AllData %>% group_by(Habitat, Transect) %>% summarise(
	SR = n_distinct(Species)
)
#now get mean, sd
SRdetailsAll<-SRbyHabAll%>% group_by(Habitat) %>% summarise(
	meanSR = mean(SR, na.rm = T),
	semSR = sd(SR, na.rm = T)/sqrt(sum(!is.na(SR)))
)
#write file
write.csv(SRbyHabAll, "AllMammSR.csv")
```


This information completes questions 1 and 2 on the second problem set.

#### Simpson's index
Let's calculate Simpson's index for our three datasets

#####Small mammals
First calculate without assuming habitat differences

```{r}
#Set up a table to receive all the information from the 3 groups:

NoHabs<-data.frame(matrix(ncol = 2, nrow = 3))
colnames(NoHabs)<-c("Group", "Simpson")
```
Now work on small mammals
```{r}
SMs<-Small%>% group_by(Species) %>% summarise(
	Total = sum(Number)
)

AllCritters<-sum(SMs$Total) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
SMs$intermediate<- ((SMs$Total)*(SMs$Total-1))/denom
Simpson<-sum(SMs$intermediate)
Group<-"Small mammals"

NoHabs$Group[1]<-Group
NoHabs$Simpson[1]<-Simpson

```

Begin by getting a separate dataset for each transect

```{r}
P1<-filter(Small, Habitat == "Pine" & Transect == 1)
P2<-filter(Small, Habitat == "Pine" & Transect == 2)
D1<-filter(Small, Habitat == "Deciduous" & Transect == 1)
D2<-filter(Small, Habitat == "Deciduous" & Transect == 2)
H1<-filter(Small, Habitat == "Hemlock" & Transect == 1)
H2<-filter(Small, Habitat == "Hemlock" & Transect == 2)
```

Could we have done that more easily w/dplyr?
```{r}
SumSmall<-Small %>% group_by(Habitat, Transect) %>% 
	summarize (
		NumIndividuals=length(Species),
		NumSpecies=n_distinct(Species)
	)
```
Now calculate Simpson's index using formula in the analysis guide.
First set up a DF to receive the results
```{r}
Table1<-data.frame(matrix(ncol = 3, nrow = 6))
colnames(Table1)<-c("Habitat", "Transect", "Simpson")
```
Start with Deciduous
```{r}
#Deciduous 1
AllCritters<-sum(D1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
D1$intermediate<- ((D1$Number)*(D1$Number-1))/denom
Simpson<-sum(D1$intermediate)
Table1$Habitat[1]<-"Deciduous"
Table1$Transect[1]<-1
Table1$Simpson[1]<-Simpson


#Deciduous 2
AllCritters<-sum(D2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
D2$intermediate<- ((D2$Number)*(D2$Number-1))/denom
Simpson<-sum(D2$intermediate)
Table1$Habitat[2]<-"Deciduous"
Table1$Transect[2]<-2
Table1$Simpson[2]<-Simpson
```
Now Hemlock
```{r}
#Hemlock 1
AllCritters<-sum(H1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
H1$intermediate<- ((H1$Number)*(H1$Number-1))/denom
Simpson<-sum(H1$intermediate)
Table1$Habitat[3]<-"Hemlock"
Table1$Transect[3]<-1
Table1$Simpson[3]<-Simpson


#Hemlock 2
AllCritters<-sum(H2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
H2$intermediate<- ((H2$Number)*(H2$Number-1))/denom
Simpson<-sum(H2$intermediate)
Table1$Habitat[4]<-"Hemlock"
Table1$Transect[4]<-2
Table1$Simpson[4]<-Simpson
```

And now Pine
```{r}
#Pine 1
AllCritters<-sum(P1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
P1$intermediate<- ((P1$Number)*(P1$Number-1))/denom
Simpson<-sum(P1$intermediate)
Table1$Habitat[5]<-"Pine"
Table1$Transect[5]<-1
Table1$Simpson[5]<-Simpson


#Pine 2
AllCritters<-sum(P2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
P2$intermediate<- ((P2$Number)*(P2$Number-1))/denom
Simpson<-sum(P2$intermediate)
Table1$Habitat[6]<-"Pine"
Table1$Transect[6]<-2
Table1$Simpson[6]<-Simpson
```
Now save result
```{r}
write.csv(Table1, "smallMammalSimpsons.csv")
```

Just need to average for each transect now.
```{r}
Tab1Simpson<-Table1 %>% group_by(Habitat) %>% summarize(
			meanSimp = mean(Simpson, na.rm = T), 
			sem_Simp=sd(Simpson)/sqrt(sum(!is.na(Simpson)))
) 
write.csv(Tab1Simpson, "small_mammal_simp_summary.csv")
```

Now can I loop through this somehow to make it faster?  I'll figure this out later, when I'm not trying to get the key...

#####Game camera data
Now work on game camera data

```{r}
GCs<-Mammals%>% group_by(Species) %>% summarise(
	Total = sum(Number)
)

AllCritters<-sum(GCs$Total) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
GCs$intermediate<- ((GCs$Total)*(GCs$Total-1))/denom
Simpson<-sum(GCs$intermediate)
Group<-"Camera"

NoHabs$Group[2]<-Group
NoHabs$Simpson[2]<-Simpson
```


Begin by getting a separate dataset for each transect

```{r}
P1<-filter(Mammals, Habitat == "Pine" & Transect == 1)
P2<-filter(Mammals, Habitat == "Pine" & Transect == 2)
D1<-filter(Mammals, Habitat == "Deciduous" & Transect == 1)
D2<-filter(Mammals, Habitat == "Deciduous" & Transect == 2)
H1<-filter(Mammals, Habitat == "Hemlock" & Transect == 1)
H2<-filter(Mammals, Habitat == "Hemlock" & Transect == 2)
```

Now calculate Simpson's index using formula in the analysis guide.
First set up a DF to receive the results
```{r}
Table2<-data.frame(matrix(ncol = 3, nrow = 6))
colnames(Table2)<-c("Habitat", "Transect", "Simpson")
```
Start with Deciduous
```{r}
#Deciduous 1
AllCritters<-sum(D1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
D1$intermediate<- ((D1$Number)*(D1$Number-1))/denom
Simpson<-sum(D1$intermediate)
Table2$Habitat[1]<-"Deciduous"
Table2$Transect[1]<-1
Table2$Simpson[1]<-Simpson


#Deciduous 2
AllCritters<-sum(D2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
D2$intermediate<- ((D2$Number)*(D2$Number-1))/denom
Simpson<-sum(D2$intermediate)
Table2$Habitat[2]<-"Deciduous"
Table2$Transect[2]<-2
Table2$Simpson[2]<-Simpson
```
Now Hemlock
```{r}
#Hemlock 1
AllCritters<-sum(H1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
H1$intermediate<- ((H1$Number)*(H1$Number-1))/denom
Simpson<-sum(H1$intermediate)
Table2$Habitat[3]<-"Hemlock"
Table2$Transect[3]<-1
Table2$Simpson[3]<-Simpson


#Hemlock 2
AllCritters<-sum(H2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
H2$intermediate<- ((H2$Number)*(H2$Number-1))/denom
Simpson<-sum(H2$intermediate)
Table2$Habitat[4]<-"Hemlock"
Table2$Transect[4]<-2
Table2$Simpson[4]<-Simpson
```

And now Pine
```{r}
#Pine 1
AllCritters<-sum(P1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
P1$intermediate<- ((P1$Number)*(P1$Number-1))/denom
Simpson<-sum(P1$intermediate)
Table2$Habitat[5]<-"Pine"
Table2$Transect[5]<-1
Table2$Simpson[5]<-Simpson


#Pine 2
AllCritters<-sum(P2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
P2$intermediate<- ((P2$Number)*(P2$Number-1))/denom
Simpson<-sum(P2$intermediate)
Table2$Habitat[6]<-"Pine"
Table2$Transect[6]<-2
Table2$Simpson[6]<-Simpson
```
Now save result
```{r}
write.csv(Table2, "GameCamSimpsons.csv")
```

Just need to average for each transect now.
```{r}
Tab2Simpson<-Table2 %>% group_by(Habitat) %>% summarize(
			meanSimp = mean(Simpson, na.rm = T), 
			sem_Simp=sd(Simpson)/sqrt(sum(!is.na(Simpson)))
) 
write.csv(Tab2Simpson, "game_cam_simp_summary.csv")
```


#####All mammals
```{r}
ALL<-AllData%>% group_by(Species) %>% summarise(
	Total = sum(Number)
)

AllCritters<-sum(ALL$Total) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
ALL$intermediate<- ((ALL$Total)*(ALL$Total-1))/denom
Simpson<-sum(ALL$intermediate)
Group<-"All mammals"

NoHabs$Group[3]<-Group
NoHabs$Simpson[3]<-Simpson
```

Begin by getting a separate dataset for each transect

```{r}
P1<-filter(AllData, Habitat == "Pine" & Transect == 1)
P2<-filter(AllData, Habitat == "Pine" & Transect == 2)
D1<-filter(AllData, Habitat == "Deciduous" & Transect == 1)
D2<-filter(AllData, Habitat == "Deciduous" & Transect == 2)
H1<-filter(AllData, Habitat == "Hemlock" & Transect == 1)
H2<-filter(AllData, Habitat == "Hemlock" & Transect == 2)
```

Now calculate Simpson's index using formula in the analysis guide.
First set up a DF to receive the results
```{r}
Table3<-data.frame(matrix(ncol = 3, nrow = 6))
colnames(Table3)<-c("Habitat", "Transect", "Simpson")
```
Start with Deciduous
```{r}
#Deciduous 1
AllCritters<-sum(D1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
D1$intermediate<- ((D1$Number)*(D1$Number-1))/denom
Simpson<-sum(D1$intermediate)
Table3$Habitat[1]<-"Deciduous"
Table3$Transect[1]<-1
Table3$Simpson[1]<-Simpson


#Deciduous 2
AllCritters<-sum(D2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
D2$intermediate<- ((D2$Number)*(D2$Number-1))/denom
Simpson<-sum(D2$intermediate)
Table3$Habitat[2]<-"Deciduous"
Table3$Transect[2]<-2
Table3$Simpson[2]<-Simpson
```
Now Hemlock
```{r}
#Hemlock 1
AllCritters<-sum(H1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
H1$intermediate<- ((H1$Number)*(H1$Number-1))/denom
Simpson<-sum(H1$intermediate)
Table3$Habitat[3]<-"Hemlock"
Table3$Transect[3]<-1
Table3$Simpson[3]<-Simpson


#Hemlock 2
AllCritters<-sum(H2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
H2$intermediate<- ((H2$Number)*(H2$Number-1))/denom
Simpson<-sum(H2$intermediate)
Table3$Habitat[4]<-"Hemlock"
Table3$Transect[4]<-2
Table3$Simpson[4]<-Simpson
```

And now Pine
```{r}
#Pine 1
AllCritters<-sum(P1$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
P1$intermediate<- ((P1$Number)*(P1$Number-1))/denom
Simpson<-sum(P1$intermediate)
Table3$Habitat[5]<-"Pine"
Table3$Transect[5]<-1
Table3$Simpson[5]<-Simpson


#Pine 2
AllCritters<-sum(P2$Number) #returns total number of critters on the transect
MostCrits<-AllCritters - 1
denom<-AllCritters*MostCrits
P2$intermediate<- ((P2$Number)*(P2$Number-1))/denom
Simpson<-sum(P2$intermediate)
Table3$Habitat[6]<-"Pine"
Table3$Transect[6]<-2
Table3$Simpson[6]<-Simpson
```
Now save result
```{r}
write.csv(Table3, "AllMammalsSimpsons.csv")
```

Just need to average for each transect now.
```{r}
Tab3Simpson<-Table3 %>% group_by(Habitat) %>% summarize(
			meanSimp = mean(Simpson, na.rm = T), 
			sem_Simp=sd(Simpson)/sqrt(sum(!is.na(Simpson)))
) 
write.csv(Tab3Simpson, "all_mammals_simp_summary.csv")
```





