---
title: "Summarizing Game Camera Data"
output: html_notebook
---

After a LOT of work processing data from the zooniverse, we end up with a file called
final_test_results.csv that contains the species IDs, habitat, transect, and the number of folks who assigned each species. We'll use this to get some summary information.

Set up the workspace and read in data
```{r}
rm(list = ls())
library(readr)
library(dplyr)
library(ggplot2)
DF <- read_csv("final_test_results.csv")
```

Now summarize list of species by subject id, transect and habitat
```{r}
newDF<-DF %>% group_by(subject_ids, transect, habitat, choice) %>% summarize(
	num_species=n_distinct(choice)
)
```
Now summarize to get frequency for each species on each transect

```{R}
SpeciesSum<-newDF %>% group_by(transect, habitat, choice) %>% summarize(
	number_species = n_distinct(choice),
	num_critters = sum(num_species, na.rm = T)
)
```
SpeciesSum is a good data frame for summarized but "raw" data

```{R}
write.csv(SpeciesSum, "summary_by_transect.csv")
```
Now summarize by habitat
```{r}
SpecByTransect <-SpeciesSum %>% group_by(habitat, transect, choice) %>% summarize(
	total = sum(num_critters, na.rm = T)
)
SpecByHabitat<-SpeciesSum %>% group_by(habitat, choice) %>% summarize(
	total = sum(num_critters, na.rm = T)
)
```
Now figure out the total number of species and individuals by transect
```{r}
ByTransect<-SpecByTransect%>%group_by(transect, habitat) %>% summarize(
	num_species = n_distinct(choice),
	total_individuals = sum(total, na.rm = T)
)
```
SpeciesByTransect has the main data I need for determining SR, Simpson's index
```{r}
write.csv(SpecByTransect, "SpeciesByTransect")
```
Now use these data elsewhere.
