---
title: 'Forest Structure 2: Mammalogy 2018'
author: "Erika"
date: "11/5/2018"
output: html_document
---

## Introduction
The students in Mammalogy are working on analyzing their forest vegetation data to help provide information about what environmental factors may play a role in determining species distributions on the Kip Tract.  In an earlier document (_"Forest Structure Error Check.Rmd"_) I showed how to clean up the data.  Now we will do questions 1 and 2 on the assignment, which are to generate a table with mean species richness, mean simpson's index for each of the 3 habitats (plus/minus SEM)

## Set up the workspace
Set some basic parameters 
```{r}
rm(list =ls()) #clear's R workspace
library(dplyr)
library(ggplot2)
```

###Step 1. Forest Diversity
Students are asked to calculate average species richness and average Simpson's index for each habitat, using each of the 2 transects within each habitat as a replicate. They are also to provide the standard error of the mean.

#### Load some data
Load the new tree data
```{r}
Trees <- read.csv("newtrees.csv")
```
#### Species Richness
Species richness is the number of species on a transect.

```{r}
MySummary<-Trees %>% group_by(Habitat, Transect.Number) %>%
	summarize(
		num_species=n_distinct(Species))
		
	
```
Now get summary info
```{r}
SR_sum_stats<-MySummary %>% group_by (Habitat) %>% 
	summarize (
		mean_species=mean(num_species, na.rm = T),
		sem_species=sd(num_species)/sqrt(sum(!is.na(num_species)))
	)
```
This information completes the first column of data asked for as Q1 on problem set 1.

#### Simpson's index
We'll look at two ways to calculate the average simpson's index for each of the two transects for each habitat so that we can get means and sem.

The first method will be that which I described in class as the way to calculate Simpson's index.

Begin by getting a separate dataset for each transect

```{r}
P1<-filter(Trees, Habitat == "Pine" & Transect.Number == 1)
P2<-filter(Trees, Habitat == "Pine" & Transect.Number == 2)
D1<-filter(Trees, Habitat == "Deciduous" & Transect.Number == 1)
D2<-filter(Trees, Habitat == "Deciduous" & Transect.Number == 2)
H1<-filter(Trees, Habitat == "Hemlock" & Transect.Number == 1)
H2<-filter(Trees, Habitat == "Hemlock" & Transect.Number == 2)
```

Could we have done that more easily w/dplyr?
```{r}
SumTrees<-Trees %>% group_by(Habitat, Transect.Number) %>% 
	summarize (
		NumTrees=length(Species),
		NumSpecies=n_distinct(Species)
	)
```
Now, for each one of the six transects, determine the number of trees in each species.

```{r}
Pine1<-P1 %>% group_by(Species) %>% summarize (
	total = length(Species)
)
AllTrees<-sum(Pine1$total) #returns total number of trees on the transect
MostTrees<-AllTrees - 1
denom<-AllTrees*MostTrees
```
This result gives me Table 1 in the analysis_guide_part_1_VEG handout for one of the transects on one of the sites.

Now calculate Simpson's index using formula in the analysis guide.

First set up a DF to receive the results
```{r}
Table1<-data.frame(matrix(ncol = 3, nrow = 6))
colnames(Table1)<-c("Habitat", "Transect", "Simpson")
```

```{r}
Pine1$intermediate<- ((Pine1$total)*(Pine1$total-1))/denom
Simpson<-sum(Pine1$intermediate)
Table1$Habitat[1]<-"Pine"
Table1$Transect[1]<-1
Table1$Simpson[1]<-Simpson
```
Pine 2
```{r}
Pine2<-P2 %>% group_by(Species) %>% summarize (
	total = length(Species)
)
AllTrees<-sum(Pine2$total) #returns total number of trees on the transect
MostTrees<-AllTrees - 1
denom<-AllTrees*MostTrees

Pine2$intermediate<- ((Pine2$total)*(Pine2$total-1))/denom
Simpson<-sum(Pine2$intermediate)
Table1$Habitat[2]<-"Pine"
Table1$Transect[2]<-2
Table1$Simpson[2]<-Simpson
```
Deciduous 1
```{r}
Decid1<-D1 %>% group_by(Species) %>% summarize (
	total = length(Species)
)
AllTrees<-sum(Decid1$total) #returns total number of trees on the transect
MostTrees<-AllTrees - 1
denom<-AllTrees*MostTrees

Decid1$intermediate<- ((Decid1$total)*(Decid1$total-1))/denom
Simpson<-sum(Decid1$intermediate)
Table1$Habitat[3]<-"Deciduous"
Table1$Transect[3]<-1
Table1$Simpson[3]<-Simpson
```
Deciduous 2

```{r}
Decid2<-D2 %>% group_by(Species) %>% summarize (
	total = length(Species)
)
AllTrees<-sum(Decid2$total) #returns total number of trees on the transect
MostTrees<-AllTrees - 1
denom<-AllTrees*MostTrees

Decid2$intermediate<- ((Decid2$total)*(Decid2$total-1))/denom
Simpson<-sum(Decid2$intermediate)
Table1$Habitat[4]<-"Deciduous"
Table1$Transect[4]<-2
Table1$Simpson[4]<-Simpson
```
Hemlock 1

```{r}
Hem1<-H1 %>% group_by(Species) %>% summarize (
	total = length(Species)
)
AllTrees<-sum(Hem1$total) #returns total number of trees on the transect
MostTrees<-AllTrees - 1
denom<-AllTrees*MostTrees

Hem1$intermediate<- ((Hem1$total)*(Hem1$total-1))/denom
Simpson<-sum(Hem1$intermediate)
Table1$Habitat[5]<-"Hemlock"
Table1$Transect[5]<-1
Table1$Simpson[5]<-Simpson
```
Hemlock 2
```{r}
Hem2<-H2 %>% group_by(Species) %>% summarize (
	total = length(Species)
)
AllTrees<-sum(Hem2$total) #returns total number of trees on the transect
MostTrees<-AllTrees - 1
denom<-AllTrees*MostTrees

Hem2$intermediate<- ((Hem2$total)*(Hem2$total-1))/denom
Simpson<-sum(Hem2$intermediate)
Table1$Habitat[6]<-"Hemlock"
Table1$Transect[6]<-2
Table1$Simpson[6]<-Simpson
```
This completes the intermediate step I need for column 2 in table 1 on the assignment.  Just need to average for each transect now.
```{r}
Tab1Simpson<-Table1 %>% group_by(Habitat) %>% summarize(
			meanSimp = mean(Simpson, na.rm = T), 
			sem_Simp=sd(Simpson)/sqrt(sum(!is.na(Simpson)))
) 
```
Now combine to one table
```{r}
CompleteTable1<-cbind(SR_sum_stats, Tab1Simpson )
write.csv(CompleteTable1, "CompleteTable1.csv")
```



Now can I loop through this somehow to make it faster?  I'll figure this out later, when I'm not trying to get the key...


