---
title: 'Forest Structure 3: Mammalogy 2018'
author: "Erika"
date: "11/7/2018"
output: html_document
---
## Introduction
As we continue solving the Mammalogy problem set generating summary data about the vegetation structure of our three habitat types, we will now step through the process of determining averages of relative density, relative basal area, relative frequency and importance value for each habitat type.  N in each is 2, values from each of the two transects in each habitat.  These steps follow those outlined in the document "PS1_SpecRich_Simpson.Rmd"

## Step 1.  Prepare the workspace and bring in some data
Let's get R ready to rock-n-roll!
```{r}
rm(list = ls()) #clear's the workspace
library(dplyr)
library(ggplot2)
```
Now load data
```{r}
Trees <- read.csv("newtrees.csv")
```
## Step 2. Forest structure analysis
Our next task is to determine the average importance values for each species in each habitat. We'll start by getting a new DF for each habitat.

```{r}
Pines<-filter(Trees, Habitat == "Pine")
Decid<-filter(Trees, Habitat == "Deciduous")
Hem<-filter(Trees, Habitat == "Hemlock")
```
Now we'll work with each data set to do the calculations according to the Analysis guide.

### A. Average Density
We start calculating average density by determining the mean point-to-individual distance for all of the trees on a single transect within a single habitat.
We will call mean point-to-individual distance MPI, mean area per plant MAP, and average density AVGDEN.

Let's begin by creating a new data frame to take all of the data

```{r}
temp_DF<-data.frame(matrix(ncol = 5, nrow = 6))
colnames(temp_DF)<-c("Habitat", "Transect", "MPI", "MAP", "AVGDEN")
```

```{r}
calcsPine <- Pines %>% group_by(Transect.Number) %>% summarise(
		MPI = mean(Distance, na.rm = T),
		MAP = MPI*MPI,
		AVGDEN = (100/MAP)
)
```
Now repeat for the other 2 habitats .
```{r}
calcsHem<-Hem %>% group_by(Transect.Number) %>% summarise(
	MPI = mean(Distance, na.rm = T),
	MAP = MPI*MPI,
	AVGDEN = (100/MAP)
)
calcsDecid<-Decid %>% group_by(Transect.Number) %>% summarise(
	MPI = mean(Distance, na.rm = T),
	MAP = MPI*MPI,
	AVGDEN = (100/MAP)
)
```
Now let's try merging the 3 DFs into one to reduce junk. Use rbind and then add
a column indicating habitat. Then we'll remove some "intermediate" DFs to clean up the workspace.
```{r}
AvgDenData<-rbind(calcsPine, calcsDecid, calcsHem)
AvgDenData$Habitat<-c("Pine", "Pine", "Decid", "Decid", "Hem", "Hem")
rm(calcsDecid, calcsHem, calcsPine, temp_DF)#removes the DFs I no longer need
```
### B. Relative density, Relative basal area, relative frequency and importance value

Now work through each transect calculating all of the steps that give you relative density, relative basal area, relative frequency and Importance Value.
#### First Transect
##### Relative Density
Get number of trees of each species on each transect
```{r}
P1<-filter(Pines, Transect.Number == 1)
TreesP1<-P1 %>% group_by(Species) %>% summarize (
	total = length(Species),
)
```
Now get total number of trees and calculate relative density and species density
```{r}
AllTreesP1<-sum(TreesP1$total)

TreesP1$RELDEN<-(TreesP1$total/AllTreesP1)*100

TreesP1$DEN<-((TreesP1$RELDEN)/100)*AvgDenData$AVGDEN[1]

#check that it is correct: all densities together should add to the average density for the transect

AvgDenData$AVGDEN[1] == sum(TreesP1$DEN)

```
##### Relative basal area
First step is that we need to calculate the basal area of each tree in the transect to area.

```{r}
P1$Radius<-P1$DBH/2
P1$Area<-pi*(P1$Radius^2)
AVGBA<- P1 %>% group_by(Species)  %>% summarise(
	AVGBA=mean(Area, na.rm = T)
) #returns class a DF with 2 columns; can't easily directly combine with TreesP1 so do this step then next line

TreesP1$AVGBA<-AVGBA$AVGBA

```
Now calculate BA and relba
```{r}
TreesP1$BA<-TreesP1$DEN*TreesP1$AVGBA
TotalBA<-sum(TreesP1$BA)
TreesP1$RELBA<-((TreesP1$BA/TotalBA)*100)
```
##### Relative frequency and Importance Value
Now frequency and relative frequency. Frequency will be a bit harder, since it requires assessing for each species which trap stations the species occurred at out of 10 possible.Let's try a loop.

```{r}
NumStations<-10
MySpecies<-levels(factor(P1$Species))
Freq<-vector()
temp<-vector()
i<-1
for (i in 1:length(MySpecies)){
	test<-filter(P1, Species == MySpecies[i])
	Stations<-unique(test$Trap.Station)
	temp[i]<-length(Stations)
	Freq[i]<-temp[i]/NumStations
}
TreesP1$FREQ<-Freq
SumFreq<-sum(TreesP1$FREQ)
TreesP1$RELFREQ<-(TreesP1$FREQ/SumFreq)*100
TreesP1$IV<-TreesP1$RELDEN+TreesP1$RELBA+TreesP1$RELFREQ
#now assign the correct transect and habitat to keep track of later
TreesP1$Habitat<-as.factor("Pine")
TreesP1$Transect.Number <- 1

```
This works, and gives me correct numbers according to my excel calculations.  I've already
found 2 human error issues on excel that are mistakes R did not make!


Now just re-run for the other 5 transects and get ready to create averages.


#### Second Transect
##### Relative Density
Get number of trees of each species on each transect
```{r}
P2<-filter(Pines, Transect.Number == 2)
TreesP2<-P2 %>% group_by(Species) %>% summarize (
	total = length(Species),
)
```
Now get total number of trees and calculate relative density and species density
```{r}
AllTreesP2<-sum(TreesP2$total)

TreesP2$RELDEN<-(TreesP2$total/AllTreesP2)*100

TreesP2$DEN<-((TreesP2$RELDEN)/100)*AvgDenData$AVGDEN[2]

#check that it is correct: all densities together should add to the average density for the transect

AvgDenData$AVGDEN[2] == sum(TreesP2$DEN)

```
##### Relative basal area
First step is that we need to calculate the basal area of each tree in the transect to area.

```{r}
P2$Radius<-P2$DBH/2
P2$Area<-pi*(P2$Radius^2)
AVGBA<- P2 %>% group_by(Species)  %>% summarise(
	AVGBA=mean(Area, na.rm = T)
) #returns class a DF with 2 columns; can't easily directly combine with TreesP2 so do this step then next line

TreesP2$AVGBA<-AVGBA$AVGBA

```
Now calculate BA and relba
```{r}
TreesP2$BA<-TreesP2$DEN*TreesP2$AVGBA
TotalBA<-sum(TreesP2$BA)
TreesP2$RELBA<-((TreesP2$BA/TotalBA)*100)
```
##### Relative frequency and Importance Value
Now frequency and relative frequency. Frequency will be a bit harder, since it requires assessing for each species which trap stations the species occurred at out of 10 possible.Let's try a loop.

```{r}
NumStations<-10
MySpecies<-levels(factor(P2$Species))
Freq<-vector()
temp<-vector()
i<-1
for (i in 1:length(MySpecies)){
	test<-filter(P2, Species == MySpecies[i])
	Stations<-unique(test$Trap.Station)
	temp[i]<-length(Stations)
	Freq[i]<-temp[i]/NumStations
}
TreesP2$FREQ<-Freq
SumFreq<-sum(TreesP2$FREQ)
TreesP2$RELFREQ<-(TreesP2$FREQ/SumFreq)*100
TreesP2$IV<-TreesP2$RELDEN+TreesP2$RELBA+TreesP2$RELFREQ
#now assign the correct transect and habitat to keep track of later
TreesP2$Habitat<-as.factor("Pine")
TreesP2$Transect.Number <- 2

```
####Third Transect
##### Relative Density
Get number of trees of each species on each transect
```{r}
D1<-filter(Decid, Transect.Number == 1)
TreesD1<-D1 %>% group_by(Species) %>% summarize (
	total = length(Species),
)
```
Now get total number of trees and calculate relative density and species density
```{r}
AllTreesD1<-sum(TreesD1$total)

TreesD1$RELDEN<-(TreesD1$total/AllTreesD1)*100

TreesD1$DEN<-((TreesD1$RELDEN)/100)*AvgDenData$AVGDEN[3]

#check that it is correct: all densities together should add to the average density for the transect

AvgDenData$AVGDEN[3] == sum(TreesD1$DEN)

```
##### Relative basal area
First step is that we need to calculate the basal area of each tree in the transect to area.

```{r}
D1$Radius<-D1$DBH/2
D1$Area<-pi*(D1$Radius^2)
AVGBA<- D1 %>% group_by(Species)  %>% summarise(
	AVGBA=mean(Area, na.rm = T)
) #returns class a DF with 2 columns; can't easily directly combine with TreesD1 so do this step then next line

TreesD1$AVGBA<-AVGBA$AVGBA

```
Now calculate BA and relba
```{r}
TreesD1$BA<-TreesD1$DEN*TreesD1$AVGBA
TotalBA<-sum(TreesD1$BA)
TreesD1$RELBA<-((TreesD1$BA/TotalBA)*100)
```
##### Relative frequency and Importance Value
Now frequency and relative frequency. Frequency will be a bit harder, since it requires assessing for each species which trap stations the species occurred at out of 10 possible.Let's try a loop.

```{r}
NumStations<-10
MySpecies<-levels(factor(D1$Species))
Freq<-vector()
temp<-vector()
i<-1
for (i in 1:length(MySpecies)){
	test<-filter(D1, Species == MySpecies[i])
	Stations<-unique(test$Trap.Station)
	temp[i]<-length(Stations)
	Freq[i]<-temp[i]/NumStations
}
TreesD1$FREQ<-Freq
SumFreq<-sum(TreesD1$FREQ)
TreesD1$RELFREQ<-(TreesD1$FREQ/SumFreq)*100
TreesD1$IV<-TreesD1$RELDEN+TreesD1$RELBA+TreesD1$RELFREQ
#now assign the correct transect and habitat to keep track of later
TreesD1$Habitat<-as.factor("Decid")
TreesD1$Transect.Number <- 1
```
#### Fourth Transect
##### Relative Density
Get number of trees of each species on each transect
```{r}
D2<-filter(Decid, Transect.Number == 2)
TreesD2<-D2 %>% group_by(Species) %>% summarize (
	total = length(Species),
)
```
Now get total number of trees and calculate relative density and species density
```{r}
AllTreesD2<-sum(TreesD2$total)

TreesD2$RELDEN<-(TreesD2$total/AllTreesD2)*100

TreesD2$DEN<-((TreesD2$RELDEN)/100)*AvgDenData$AVGDEN[4]

#check that it is correct: all densities together should add to the average density for the transect

AvgDenData$AVGDEN[4] == sum(TreesD2$DEN)

```
##### Relative basal area
First step is that we need to calculate the basal area of each tree in the transect to area.

```{r}
D2$Radius<-D2$DBH/2
D2$Area<-pi*(D2$Radius^2)
AVGBA<- D2 %>% group_by(Species)  %>% summarise(
	AVGBA=mean(Area, na.rm = T)
) #returns class a DF with 2 columns; can't easily directly combine with TreesD2 so do this step then next line

TreesD2$AVGBA<-AVGBA$AVGBA

```
Now calculate BA and relba
```{r}
TreesD2$BA<-TreesD2$DEN*TreesD2$AVGBA
TotalBA<-sum(TreesD2$BA)
TreesD2$RELBA<-((TreesD2$BA/TotalBA)*100)
```
##### Relative frequency and Importance Value
Now frequency and relative frequency. Frequency will be a bit harder, since it requires assessing for each species which trap stations the species occurred at out of 10 possible.Let's try a loop.

```{r}
NumStations<-10
MySpecies<-levels(factor(D2$Species))
Freq<-vector()
temp<-vector()
i<-1
for (i in 1:length(MySpecies)){
	test<-filter(D2, Species == MySpecies[i])
	Stations<-unique(test$Trap.Station)
	temp[i]<-length(Stations)
	Freq[i]<-temp[i]/NumStations
}
TreesD2$FREQ<-Freq
SumFreq<-sum(TreesD2$FREQ)
TreesD2$RELFREQ<-(TreesD2$FREQ/SumFreq)*100
TreesD2$IV<-TreesD2$RELDEN+TreesD2$RELBA+TreesD2$RELFREQ
#now assign the correct transect and habitat to keep track of later
TreesD2$Habitat<-as.factor("Decid")
TreesD2$Transect.Number <- 2
```
#### Fifth transect
##### Relative Density
Get number of trees of each species on each transect
```{r}
H1<-filter(Hem, Transect.Number == 1)
TreesH1<-H1 %>% group_by(Species) %>% summarize (
	total = length(Species),
)
```
Now get total number of trees and calculate relative density and species density
```{r}
AllTreesH1<-sum(TreesH1$total)

TreesH1$RELDEN<-(TreesH1$total/AllTreesH1)*100

TreesH1$DEN<-((TreesH1$RELDEN)/100)*AvgDenData$AVGDEN[5]

#check that it is correct: all densities together should add to the average density for the transect

AvgDenData$AVGDEN[5] == sum(TreesH1$DEN)

```
##### Relative basal area
First step is that we need to calculate the basal area of each tree in the transect to area.

```{r}
H1$Radius<-H1$DBH/2
H1$Area<-pi*(H1$Radius^2)
AVGBA<- H1 %>% group_by(Species)  %>% summarise(
	AVGBA=mean(Area, na.rm = T)
) #returns class a DF with 2 columns; can't easily directly combine with TreesH1 so do this step then next line

TreesH1$AVGBA<-AVGBA$AVGBA

```
Now calculate BA and relba
```{r}
TreesH1$BA<-TreesH1$DEN*TreesH1$AVGBA
TotalBA<-sum(TreesH1$BA)
TreesH1$RELBA<-((TreesH1$BA/TotalBA)*100)
```
##### Relative frequency and Importance Value
Now frequency and relative frequency. Frequency will be a bit harder, since it requires assessing for each species which trap stations the species occurred at out of 10 possible.Let's try a loop.

```{r}
NumStations<-10
MySpecies<-levels(factor(H1$Species))
Freq<-vector()
temp<-vector()
i<-1
for (i in 1:length(MySpecies)){
	test<-filter(H1, Species == MySpecies[i])
	Stations<-unique(test$Trap.Station)
	temp[i]<-length(Stations)
	Freq[i]<-temp[i]/NumStations
}
TreesH1$FREQ<-Freq
SumFreq<-sum(TreesH1$FREQ)
TreesH1$RELFREQ<-(TreesH1$FREQ/SumFreq)*100
TreesH1$IV<-TreesH1$RELDEN+TreesH1$RELBA+TreesH1$RELFREQ
#now assign the correct transect and habitat to keep track of later
TreesH1$Habitat<-as.factor("Hem")
TreesH1$Transect.Number <- 1
```
#### Sixth Transect
##### Relative Density
Get number of trees of each species on each transect
```{r}
H2<-filter(Hem, Transect.Number == 2)
TreesH2<-H2 %>% group_by(Species) %>% summarize (
	total = length(Species),
)
```
Now get total number of trees and calculate relative density and species density
```{r}
AllTreesH2<-sum(TreesH2$total)

TreesH2$RELDEN<-(TreesH2$total/AllTreesH2)*100

TreesH2$DEN<-((TreesH2$RELDEN)/100)*AvgDenData$AVGDEN[6]

#check that it is correct: all densities together should add to the average density for the transect

AvgDenData$AVGDEN[6] == sum(TreesH2$DEN)

```
##### Relative basal area
First step is that we need to calculate the basal area of each tree in the transect to area.

```{r}
H2$Radius<-H2$DBH/2
H2$Area<-pi*(H2$Radius^2)
AVGBA<- H2 %>% group_by(Species)  %>% summarise(
	AVGBA=mean(Area, na.rm = T)
) #returns class a DF with 2 columns; can't easily directly combine with TreesH2 so do this step then next line

TreesH2$AVGBA<-AVGBA$AVGBA

```
Now calculate BA and relba
```{r}
TreesH2$BA<-TreesH2$DEN*TreesH2$AVGBA
TotalBA<-sum(TreesH2$BA)
TreesH2$RELBA<-((TreesH2$BA/TotalBA)*100)
```
##### Relative frequency and Importance Value
Now frequency and relative frequency. Frequency will be a bit harder, since it requires assessing for each species which trap stations the species occurred at out of 10 possible.Let's try a loop.

```{r}
NumStations<-10
MySpecies<-levels(factor(H2$Species))
Freq<-vector()
temp<-vector()
i<-1
for (i in 1:length(MySpecies)){
	test<-filter(H2, Species == MySpecies[i])
	Stations<-unique(test$Trap.Station)
	temp[i]<-length(Stations)
	Freq[i]<-temp[i]/NumStations
}
TreesH2$FREQ<-Freq
SumFreq<-sum(TreesH2$FREQ)
TreesH2$RELFREQ<-(TreesH2$FREQ/SumFreq)*100
TreesH2$IV<-TreesH2$RELDEN+TreesH2$RELBA+TreesH2$RELFREQ
#now assign the correct transect and habitat to keep track of later
TreesH2$Habitat<-as.factor("Hem")
TreesH2$Transect.Number <- 2
```

Now squish into one DF to make summary easier and to clean up workspace

```{r}
AllData<-rbind(TreesD1,TreesD2, TreesH1, TreesH2, TreesP1, TreesP2)
#now get rid of some stuff
rm(AVGBA, D1,D2, Decid, H1, H2, Hem, P1, P2, Pines, test, Trees)
```

#now let's summarize 
```{r}
SummaryData<-AllData %>% group_by(Habitat, Species)%>% summarise(
	N = length(Species),
		avgRELDEN = mean(RELDEN, na.rm = T),
		semRELDEN = sd(RELDEN)/sqrt(sum(!is.na(RELDEN))),
		avgRELBA = mean(RELBA, na.rm = T),
		semRELBA = sd(RELBA)/sqrt(sum(!is.na(RELBA))),
		avgRELFREQ = mean(RELFREQ, na.rm = T),
		semRELFREQ = sd(RELFREQ)/sqrt(sum(!is.na(RELFREQ))),
		avgIV = mean (IV, na.rm = T),
		semIV = sd(IV)/sqrt(sum(!is.na(IV)))
	)
#note: if you use the same variable name for the new DF (in this case SummaryData)
#as you have in the first DF (in this case AllData), you will not get the sem calculations.

#this summary table produces the information required for the long table in the problem set

#write results
write.csv(SummaryData, "longPS1table.csv")

```

Now create plot required in PS1

```{r}
#really nice plot with facet wrap and error bars
ggplot(SummaryData, aes(x = Species, y = avgIV, fill = Habitat))+
	geom_col(position = "dodge")+
	coord_flip()+
	geom_errorbar(aes(ymin = avgIV - semIV,
				   ymax = avgIV + semIV), width = 0.1)+
	facet_wrap(~Habitat)+
	labs(x = "Species", y = "Average Importance Value")

ggsave("PS1figure.png", device = "png")

#less clear plot without facet wrap
ggplot(SummaryData, aes(x = Species, y = avgIV, fill = Habitat))+
	geom_col(position = "dodge")+
	coord_flip()+
	geom_errorbar(aes(ymin = avgIV - semIV,
				   ymax = avgIV + semIV))+
	labs(x = "Species", y = "Average Importance Value")
```




### Try a loop
First, concatenate Habitat and transect to give a unique name for each transect.
```{r}
#Trees<-mutate(Trees, TranNum = paste0(Habitat, Transect.Number))#create new var
```
Next, get variable with list of transects
```{r}
#Transects<-unique(Trees$TranNum)
```
Now I can do a loop through each transect.  I will make a new DF for each transect and then use them later.

```{r}
#for(i in 1: length(Transects)){
	#DF<-filter(Trees, TranNum == Transects[i]) #creates DF with first transect
	#Species<-DF %>% group_by(Species) %>% summarize (
	#	total= length(Species)
	#)
	#AllTrees<-sum(Species$total)
#}	
```




