---
title: "Ground Cover estimates"
output: html_notebook
---
## Introduction
This markdown document explains how to generate a table showing the average 
percent bare ground, % vascular cover and % bryophyte cover, as well as litter depth
for each of the forest habitats, as well as their respective means.

## Step 1. Prep the workspace and turn on desired packages
```{r}
rm(list = ls())
library(dplyr)
library(ggplot2)
library(reshape2)
```
## Step 2. Load the data and prep it
Load the data
```{r}
library(readr)
DF <- read_csv("newground.csv")
#rename some variables
names(DF)<-c("Row", "Date", "Habitat", "TranNum", "Team", "Station", "Quad", "Measurement", "Value", 
		   "Comments")
```
Now concatenate the habitat and transect number columns to give specific columns
```{r}
DF<-mutate(DF, HabNum = paste0(Habitat,"_",TranNum))#create new var
```
#Step 3 Summarize data
First summarize by transect (10 values per transect)
```{r}
SummaryData<-DF %>% group_by(Habitat, HabNum, Measurement)%>% summarise(
	N = length(Value),
	meanVal= mean(Value, na.rm = T),
	semVal = sd(Value, na.rm = T)/sqrt(N)
)


```
Now summarize by habitat type
```{r}
PS1Table3<-SummaryData %>% group_by(Habitat, Measurement) %>% summarise(
		N = length(meanVal),
	meanVal1= mean(meanVal, na.rm = T),
	semVal1 = sd(meanVal, na.rm = T)/sqrt(N)
)
```
Now save data
```{r}
write.csv(PS1Table3, "PS1Table3.csv")
```
Now get a nice plot

```{r}
ggplot(PS1Table3, aes(x = Habitat, y = meanVal1, fill = Measurement))+
geom_col(position = "dodge")+
	coord_flip()+
	geom_errorbar(aes(ymin = meanVal1 - semVal1,
				   ymax = meanVal1 + semVal1), width = 0.1)+
	facet_wrap(~Measurement)+
	labs(x = "Habitat", y = "Mean value")

ggsave("PS3figure.png", device = "png")
```


