---
title: "Using adehabitatHR, adehabitatLT and sp packages to measure animal home range"
output: html_notebook
---

## Introduction
Animal home ranges are the areas that animals use to carry out their normal day-to-day activities such as sleeping, feeding, and mating.  There at two general classes of home range estimators: *Minimum Convex Polygon* (MCP) and *Kernel Density Estimator* (KDE) home range estimates.  

In the first part of this tutorial, we will examine how to load data and let R know what it needs to know about the data.

In the second part we will calculate the MCP home ranges for Cape Buffalo in South Africa.

In the third part we will calculate KDE home ranges for the same Cape Buffalo.

In the final part, we will compare the MCP to KDE estimators to see the pros and cons of each.

So, let's begin!

## Part 1 - Prepping the R workspace and the data
I like to begin any session in R with setting up a few features of the workspace and loading any packages that I will need.

```{r}
rm(list=ls()) #clear's the environment

#now install pacman package to check on what packages that you need are already installed and install them if not
if(!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, adehabitatHR, adehabitatLT, sp, ggplot2, magrittr, maptools) #checks to see if packages are installed and installs them if not

#now turn the packages on
library(dplyr)
library(tidyr)
library(adehabitatHR)
library(adehabitatLT)
library(sp)
library(rgdal)#won't load on rstudio server; used to apply datum, export .shp
library(ggplot2)
library(magrittr)
library(maptools)
library(tidyr)
library(sf)
library(here)
```

Ok, now that we've got stuff cleared, turned on, and ready to go, let's load some data.

```{r}
CapeBuffalo <- read.csv(here("Mammalogy18", "SpaceUse", "Data", "CapeBuffalo", "Kruger African Buffalo, GPS tracking, South Africa.csv"))
```
Now let's look at the data and see how they are organized
```{r}
structure(CapeBuffalo) #examines the structure of the dataset
```
Now let's plot the data

In order to work with the home range data, we need first to subset just the location and animal identification information and then make sure R recognizes that the location data are spatial.
```{r}
plot(CapeBuffalo$'utm.easting', CapeBuffalo$'utm.northing', asp = 1)
```

```{r}
#subset DF to get just animal name, X and Y coords

Data<-CapeBuffalo %>% dplyr::select('utm.easting', 'utm.northing', `individual.local.identifier`)

str(Data)
```
Now let's tell R that the utm coordinates represent spatial and not just numeric information.
```{r}
sp::coordinates(Data)<-Data[,c("utm.easting", "utm.northing")]

```
Ok! Our data are ready, so we can move on to Part 2!

## Part 2 - Calculate the MCP home ranges for the buffalo.

First, let's see how many buffalo are in our dataset by examining the 'individual.local.identifer (aka Name).  To do so, we need to change the names from character variables to factor variables.

```{r}
#change from character to factor
Data$individual.local.identifier<-as.factor(Data$individual.local.identifier)

#now see how many levels of the factor there are
levels(Data$individual.local.identifier)
```
Looks like there are 6 buffalo in our data set.

Second, we can calculate the MCP estimates for each buffalo.

Often, when we estimate MCP or any other home range, we like to examine the home range size with a variety of the whole data set.

```{r}

mcp_est100<-mcp(Data[,3], percent=100, unin = "m", unout="km2") #100% of points used
mcp_est90<-mcp(Data[,3], percent=90, unin= 'm', unout = 'km2')#90 % of points used
mcp_est80<-mcp(Data[,3], percent=80, unin= 'm', unout = 'km2')#80% of points used
mcp_est70<-mcp(Data[,3], percent=70, unin= 'm', unout = 'km2')#70% of points used

#now plot the results

plot(mcp_est100, col = 'purple')
plot(mcp_est90, add = TRUE, col = 'green')
plot(mcp_est80, add = T, col = 'red')
plot(mcp_est70, add = T, col = 'blue')
plot(Data, cex = 0.5, pch = 19, col = 'orange', add = T)
```
Now use the sf package to write the shapefiles. 

```{r}
sf::st_write(as(mcp_est100, "sf"), here("Mammalogy22", "Data", "Shapefiles", "MCP100.shp"))

sf::st_write(as(mcp_est90, "sf"), here("Mammalogy22", "Data", "Shapefiles", "MCP90.shp"))

sf::st_write(as(mcp_est80, "sf"), here("Mammalogy22", "Data", "Shapefiles", "MCP80.shp"))

sf::st_write(as(mcp_est70, "sf"), here("Mammalogy22", "Data", "Shapefiles", "MCP70.shp"))

```
Now let's get the areas of the home ranges

```{r}
MCP_areas<-as.data.frame(mcp_est100)
MCP_areas
```
The result is the size, in area by square kilometers since that is what was specified by "unout", of the home range for each of the 6 animals.

## Part 3 - Calculate the KDE home ranges for the buffalo.

Ok, now let's calculate the Kernel Density Estimator for each of the buffalo.  Remember that the 3rd column of Data has the buffalo names.
```{r}
kud<-kernelUD(Data[,3], h = 'href')
kud
image(kud)#displays the results
```

Now let's estimate the home range from the Utility Distribution

```{r}
homeranges<-getverticeshr(kud, unin = "m", unout="km2")
class(homeranges)
plot(homeranges, col = 1:6)
```
Now let's export the KDE home ranges as a shapefile
```{r}
#requires sf

sf::st_write(as(homeranges, "sf"), here("Mammalogy22","Data", "Shapefiles","KDEhomeranges.shp"))

```


Now let's get the homerange areas
```{r}
KDE_areas<-as.data.frame(homeranges)
KDE_areas
```
These areas are in square kilomters.





## Part 4 - Compare the size of home ranges calculated by MCP and KDE

Start by combining all the data into a single data frame

```{r}
HRS<-data.frame(MCP_areas$id, MCP_areas$area, KDE_areas$area)
```
Now fix the column names to avoid confusion
```{r}
colnames(HRS)<-c("Name", "MCP", "KDE")
HRS
```
Now we need to re-arrange the data set so that there is a column called "Name", a column called "Type", and a column called "Area".  To do so, we'll use the package called **tidyr**

```{r}
library(tidyr)
HRS_long<-tidyr::gather(HRS, Type, Area, MCP:KDE, factor_key =T)
HRS_long
```
We did that to make it easer to plot the data.  Now let's make a plot!

```{r}
ggplot(HRS_long, aes(Name, Area, fill = Type))+
   geom_col(stat = identity, position = 'dodge')
ggsave("MCPvsKDE.png", device = "png")
```
Can we make any generalizations about MCP vs. KDE estimates based on these results?

## Part 5 - Mapping the data on an interesting background
Though we can see the data in the drawings above, it would be nice to see them displayed relative to some meaningful background.  

The first step is to convert the polygon shapefiles into a dataframe and just plot them:

```{r}
MCPs<-fortify(mcp_est100)
g<-ggplot(MCPs, aes(x = long, y = lat, fill = id, group = group))+
   geom_polygon((alpha = 0.4)+
                   ggthemes::scale_fill_gdocs()+
                   coord_equal()+
                   theme_void())
g
g+facet_wrap(~id)
```

Next, we will export the MCP polygons and KDE polygons as shapefile (.shp) to use in ArcGIS


```{r}
writeSpatialShape(MCPs,BuffaloMCPs, factor2char = T )

```


